---
title: "p8105_hw3_co2554"
author: "Camille Okonwko"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#### installing the "devtools" package in the console
```{r devtool_install, eval = FALSE}
devtools::install_github("p8105/p8105.datasets", force = TRUE)
```

### loading up the necessary datasets from the p8105.datasets library, and tidyverse
```{r package load, message = FALSE}
library(p8105.datasets)
data("instacart")
data("brfss_smart2010")
library(tidyverse)
library(viridis)
library(tidyr)
library(knitr)
```

### Problem 1: InstaCart

The Instacart dataset illustrates the online grocery orders of their application users. The dataset contains `r nrow(instacart)` observations of `r length(unique(instacart[["user_id"]]))` unique user ID's, with each row representing an item from an order. There is one order per user in this dataset. 

There are `r ncol(instacart)` variables in this dataset, including:

* `order_id`: order identifier

* `product_id`: product identifier

* `add_to_cart_order`: order in which each product was added to cart

* `reordered`: 1 if this prodcut has been ordered by this user in the past, 0 otherwise

* `user_id`: customer identifier

* `eval_set`: which evaluation set this order belongs in (Note that the data for use in this class is exclusively from the “train” eval_set)

* `order_number`: the order sequence number for this user (1=first, n=nth)

* `order_dow`: the day of the week on which the order was placed

* `order_hour_of_day`: the hour of the day on which the order was placed

* `days_since_prior_order`: days since the last order, capped at 30, NA if order_number=1

* `product_name`: name of the product

* `aisle_id`: aisle identifier

* `department_id`: department identifier

* `aisle`: the name of the aisle

* `department`: the name of the department

There are `r length(unique(instacart[["aisle"]]))` aisles in this dataset. The aisles that have the most ordered items are the `r paste(names(head(sort(table(instacart$aisle), decreasing = TRUE), 5)), collapse = ', ')` aisles, respectively. 


The following plot will illustrate the number of items ordered in each aisle: 

```{r ggp_aisle, message = FALSE}
ggp_aisle_plot =
  instacart |>
  group_by(department, aisle) |>
  summarize(total_items_ordered = sum(add_to_cart_order)) |>
  filter(total_items_ordered > 10000) |>
  arrange(department, desc(total_items_ordered)) |>
  ggplot(aes(x = reorder(aisle, -total_items_ordered), y = total_items_ordered, fill = department)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Number of Items Ordered in Each Aisle (Aisles with >10,000 items) - Grouped by Department",
    x = "Aisle",
    y = "Number of Items Ordered",
    fill = "Department"
  ) +
  scale_y_continuous(labels = scales::comma_format(scale = 1e-3)) +
  scale_x_discrete() +
  scale_fill_viridis_d(direction = -1) +
  scale_color_viridis_d(direction = -1)
```


This table includes the three most popular items in each of the aisles "baking ingredients", "dog food care:, and "packaged vegetable fruits"

```{r popular_items, message = FALSE}
popular_items_table =
  instacart |>
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle, product_name) |>
  summarize(order_count = sum(add_to_cart_order)) |>
  group_by(aisle) |>
  do(top_n(., 3, wt = order_count))

kable(popular_items_table, 
      col.names = c("Aisle", "Product Name", "Order Count"),
      align = "c")
```


This next table shows the average hour of the day Pink Lady Apples and Coffee Ice Cream are ordered.

```{r apples_coffee, message = FALSE}
apples_coffee_table =
  instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) |>
  mutate(
    product_name = factor(product_name, levels = c("Pink Lady Apples", "Coffee Ice Cream")),
    order_dow = factor(order_dow, levels = 0:6, labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
  ) |> 
  pivot_wider(names_from = order_dow, values_from = mean_hour)

kable(apples_coffee_table, 
      col.names = c("Product", "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"),
      align = "c",
      digits =2)
```



### Problem 2: BRFSS SMART 2010


 
#### loading up the BRFSS dataset from the p8105.datasets package:

```{r}

brfss_smart2010 = 
  brfss_smart2010 |> 
  as_tibble()

```

#### Let's do some data cleaning.


First, we'll format the data to include more appropriate data names. After the variable names are cleaned, we'll focus in on the "Overall Health" `topic` with the `response` "Poor", "Fair", "Good", "Very Good", and "Excellent.

```{r brfss_cleaned, message = FALSE}
brfss_smart2010_df =
brfss_smart2010 |> 
  janitor::clean_names() |> 
   filter(topic == "Overall Health" & 
         response %in% c("Excellent", "Very good", "Good", "Fair", "Poor")) |> 
  mutate(response = factor(response, levels = c("Poor", "Fair", "Good", "Very Good", "Excellent"), ordered = TRUE))
```


* In 2002, `r brfss_smart2010_df |> filter(year == 2002) |>  group_by(locationabbr) |>  summarise(total_observations = n()) |> filter(total_observations >= 7) |> count()` states were observed at 7 or more locations. In 2010 `r brfss_smart2010_df |> filter(year == 2010) |>  group_by(locationabbr) |>  summarise(total_observations = n()) |> filter(total_observations >= 7) |> count()` states were observed at 7 or more locations as well.










